---
title: "Introduction to RBPSpecificity"
author: "Soon Yi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RBPSpecificity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

**RBPSpecificity** provides tools to analyze RNA-Binding Protein (RBP) binding 
specificities from high-throughput sequencing data. This package provides tools for:

- Calculating **Inherent Specificity (IS)** from k-mer enrichment data
- Calculating **Mutational Sensitivity (MS)** for top motifs
- Visualizing specificity distributions and sensitivity profiles
- Computing enrichment scores from CLIP peak data

# Introduction

## Background

The package implements two key metrics for characterizing RBP binding behavior:

1. **Inherent Specificity (IS)**: Quantifies how specifically an RBP binds its 
   top motif relative to all other motifs. Higher IS indicates more specific binding.
2. **Mutational Sensitivity (MS)**: Measures how much the binding affinity 
   changes upon single nucleotide variations in the motif. Higher MS indicates
   greater sensitivity to sequence changes.

These metrics enable quantitative comparison of RBP binding properties across
different experimental conditions or between different RBPs.

## Reference

For a detailed description of these metrics and their biological applications, 
please refer to:

> Yi S, Singh SS, Ye X, Krishna R, Jankowsky E, Luna JM. (2025). 
> *Inherent Specificity and Mutational Sensitivity as Quantitative Metrics for RBP Binding.* 
> bioRxiv. https://www.biorxiv.org/content/10.1101/2025.03.28.646018v2

## Sample Data

The sample data included in this package were obtained from the ENCODE portal 
(Luo et al., 2020).

### RBNS Data

RNA Bind-n-Seq (RBNS) datasets provide in vitro RNA-RBP affinity measurements. 
For each K-mer enrichment dataset, the RBP concentration generating the highest 
R-value was selected as the representative sample. The collected RBNS R scores 
were feature-scaled to [1, e] followed by natural log transformation to normalize 
to a 0-to-1 scale.

### eCLIP Data

eCLIP datasets provide in vivo RBP binding site information. Peak coordinates 
were extended 25 nucleotides upstream from the 5'-end. Motif enrichment was 
calculated by counting K-mer appearances across extended peaks and subtracting 
background counts from randomly shifted genomic regions (up to 500 nt away). 
Background generation was repeated 100 times for robust averaging.

> Luo Y, Hitz BC, Gabdank I, et al. (2020). New developments on the Encyclopedia 
> of DNA Elements (ENCODE) data portal. *Nucleic Acids Res.* 48, D882–D889.

# Setup

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RBPSpecificity")
```

```{r load-package}
library(RBPSpecificity)
library(ggplot2)
```

# Pre-computed Enrichment Data (RBNS)

This workflow demonstrates how to calculate IS and MS from pre-computed 
enrichment scores, such as those from RBNS (RNA Bind-n-Seq) experiments.

## Loading RBNS Data

The sample data contains normalized 5-mer enrichment scores for two RBPs: 
HNRNPC (high specificity) and EIF4G2 (low specificity).

```{r load-rbns}
# Load sample RBNS data from package
rbns_file <- system.file("extdata", "RBNS_normalized_5mer.csv", 
                          package = "RBPSpecificity")
rbns_data <- read.csv(rbns_file, stringsAsFactors = FALSE)

# Preview the data
head(rbns_data)
```

## Preparing Data for Analysis

The package expects a data.frame with `MOTIF` and `Score` columns:

```{r prep-hnrnpc}
# Prepare HNRNPC data
hnrnpc_rbns <- data.frame(
  MOTIF = rbns_data$Motif,
  Score = rbns_data$HNRNPC
)

# Prepare EIF4G2 data
eif4g2_rbns <- data.frame(
  MOTIF = rbns_data$Motif,
  Score = rbns_data$EIF4G2
)
```

## Calculating Inherent Specificity (IS)

IS is calculated as the ratio of the top motif's score to the median score:

$$IS = \frac{Score_{top}}{Score_{median}}$$

```{r calc-is}
# Calculate IS for top motif (default)
is_hnrnpc <- returnIS(hnrnpc_rbns)
is_eif4g2 <- returnIS(eif4g2_rbns)

message("HNRNPC IS: ", round(is_hnrnpc, 2))
message("EIF4G2 IS: ", round(is_eif4g2, 2))
```

You can also calculate IS for all motifs:

```{r calc-is-all}
is_all <- returnIS(hnrnpc_rbns, return_type = "all")
head(is_all[order(-is_all$IS), ], 10)
```

## Calculating Mutational Sensitivity (MS)

MS quantifies how sensitive an RBP is to single nucleotide changes in its 
preferred motif.

```{r calc-ms}
# Get MS as a matrix (shows sensitivity at each position)
ms_matrix <- returnMS(hnrnpc_rbns, output_type = "matrix")
print(ms_matrix)

# Get MS as a single number (average sensitivity)
ms_hnrnpc <- returnMS(hnrnpc_rbns, output_type = "number")
ms_eif4g2 <- returnMS(eif4g2_rbns, output_type = "number")

message("HNRNPC MS: ", round(ms_hnrnpc, 4))
message("EIF4G2 MS: ", round(ms_eif4g2, 4))
```

## Visualization

### IS Distribution Plot

The `plotIS()` function displays the distribution of k-mer scores, highlighting 
the top motif and its IS value:

```{r plot-is-hnrnpc, fig.cap="Distribution of k-mer scores for HNRNPC (high specificity)."}
plotIS(hnrnpc_rbns)
```

You can adjust the number of histogram bins with the `bins` parameter:

```{r plot-is-eif4g2, fig.cap="Distribution of k-mer scores for EIF4G2 (low specificity) with 30 bins."}
plotIS(eif4g2_rbns, bins = 30)
```

### MS Profile Plot

The `plotMS()` function visualizes how sensitive binding is to mutations at each 
position. Here we compare MS profiles for both RBPs:

```{r plot-ms-hnrnpc, fig.cap="Mutational sensitivity profile for HNRNPC (high specificity RBP)."}
plotMS(hnrnpc_rbns)
```

```{r plot-ms-eif4g2, fig.cap="Mutational sensitivity profile for EIF4G2 (low specificity RBP)."}
plotMS(eif4g2_rbns)
```

# De Novo Enrichment from Peaks (eCLIP)

This workflow demonstrates how to calculate enrichment scores directly from 
peak data using the `motifEnrichment()` function.

> **Note**: While this example uses eCLIP-derived peaks, `motifEnrichment()` 
> works with any peak data in BED format (e.g., iCLIP, PAR-CLIP, or other 
> ChIP-seq derived peaks).

> **Requirement**: This workflow requires `BSgenome.Hsapiens.UCSC.hg38` to be installed.

## Loading Peak Data

```{r load-eclip}
# Load sample eCLIP peaks (BED format) from package
peaks_file <- system.file("extdata", "eCLIP_Peaks_HNRNPC.bed", 
                           package = "RBPSpecificity")
hnrnpc_peaks <- read.table(peaks_file, header = FALSE)
colnames(hnrnpc_peaks) <- c("chr", "start", "end", "name", "score", "strand",
                            "signalValue", "pValue", "qValue", "peak")

message("Loaded ", nrow(hnrnpc_peaks), " HNRNPC eCLIP peaks.")
```

## Running motifEnrichment Pipeline

In addition to standard input variables, `motifEnrichment()` provides several 
parameters for customization:

| Parameter | Description |
|-----------|-------------|
| `extension` | Number of base pairs to extend each peak from its 5'-end (upstream) |
| `Bkg_number` | Number of background iterations for averaging (higher = more robust) |
| `Bkg_dist` | Minimum distance (bp) to shift peaks when generating background regions |
| `max_shift_dist` | Maximum distance (bp) for shifting peaks in background generation |
| `log_transform` | If TRUE, applies log transformation after min-max scaling to [1, e] |

The `log_transform` parameter controls the final score transformation:
- When `TRUE` (default): Scores are first scaled to [1, e] using min-max 
  normalization, then the natural logarithm is applied
- When `FALSE`: Only the specified `normalization_method` is applied directly

```{r run-enrichment, eval=FALSE}
# NOTE: This requires BSgenome.Hsapiens.UCSC.hg38
# Bkg_number is set low (10) for demo; use 100+ for real analysis

hnrnpc_enrichment <- motifEnrichment(
  peak_data = hnrnpc_peaks,
  species_or_build = "hg38",
  K = 5,
  extension = 25,
  Bkg_number = 10,
  Bkg_dist = 500,
  max_shift_dist = 1000,
  log_transform = TRUE
)

# For comparison, also run on EIF4G2 peaks
eif4g2_file <- system.file("extdata", "eCLIP_Peaks_EIF4G2.bed", 
                            package = "RBPSpecificity")
eif4g2_peaks <- read.table(eif4g2_file, header = FALSE)
colnames(eif4g2_peaks) <- c("chr", "start", "end", "name", "score", "strand",
                             "signalValue", "pValue", "qValue", "peak")

eif4g2_enrichment <- motifEnrichment(
  peak_data = eif4g2_peaks,
  species_or_build = "hg38",
  K = 5,
  extension = 25,
  Bkg_number = 10,
  Bkg_dist = 500,
  max_shift_dist = 1000,
  log_transform = TRUE
)

# View top enriched motifs for each RBP
head(hnrnpc_enrichment[order(-hnrnpc_enrichment$Score), ], 10)
head(eif4g2_enrichment[order(-eif4g2_enrichment$Score), ], 10)
```


## Calculating Metrics from Enrichment

```{r eclip-metrics, eval=FALSE}
# Calculate IS and MS from enrichment data
is_eclip_hnrnpc <- returnIS(hnrnpc_enrichment)
ms_eclip_hnrnpc <- returnMS(hnrnpc_enrichment, output_type = "number")

is_eclip_eif4g2 <- returnIS(eif4g2_enrichment)
ms_eclip_eif4g2 <- returnMS(eif4g2_enrichment, output_type = "number")

message("HNRNPC eCLIP - IS: ", round(is_eclip_hnrnpc, 2), ", MS: ", round(ms_eclip_hnrnpc, 4))
message("EIF4G2 eCLIP - IS: ", round(is_eclip_eif4g2, 2), ", MS: ", round(ms_eclip_eif4g2, 4))
```

## Visualization

```{r eclip-plots, eval=FALSE}
# IS Distribution plots
plotIS(hnrnpc_enrichment)
plotIS(eif4g2_enrichment)

# MS Profile plots
plotMS(hnrnpc_enrichment)
plotMS(eif4g2_enrichment)
```

# Comparing Multiple RBPs

You can loop through multiple RBPs to build a comparison table:

```{r multi-rbp}
# Get list of RBPs from column names
rbps <- colnames(rbns_data)[-1]  # Exclude 'Motif'

# Build results table
results <- data.frame(
  RBP = rbps,
  IS = NA_real_,
  MS = NA_real_
)

for (i in seq_along(rbps)) {
  temp_df <- data.frame(
    MOTIF = rbns_data$Motif,
    Score = rbns_data[[rbps[i]]]
  )
  results$IS[i] <- returnIS(temp_df)
  results$MS[i] <- returnMS(temp_df, output_type = "number")
}

print(results)
```

# Comparing RBNS vs eCLIP

A key application is comparing **inherent specificity** (from RBNS) with 
**apparent specificity** (from eCLIP):

```{r comparison, eval=FALSE}
comparison <- data.frame(
  Source = c("RBNS (Inherent)", "eCLIP (Apparent)"),
  IS = c(is_hnrnpc, is_eclip),
  MS = c(ms_hnrnpc, ms_eclip)
)
print(comparison)
```

# Session Info

```{r session}
sessionInfo()
```
