---
title: "Introduction to RBPSpecificity"
author: "Soon Yi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RBPSpecificity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

**RBPSpecificity** provides tools to analyze RNA-Binding Protein (RBP) binding 
specificities from high-throughput sequencing data. This package provides tools for:

- Calculating **Inherent Specificity (IS)** from k-mer enrichment data
- Calculating **Variation Sensitivity (VS)** for top motifs
- Visualizing specificity distributions and sensitivity profiles
- Computing enrichment scores from CLIP peak data

# Introduction

## Background

The package implements two key metrics for characterizing RBP binding behavior:

1. **Inherent Specificity (IS)**: Quantifies how specifically an RBP binds its 
   top motif relative to all other motifs. Higher IS indicates more specific binding.
2. **Variation Sensitivity (VS)**: Measures how much the binding affinity 
   changes upon single nucleotide variations in the motif. Higher VS indicates
   greater sensitivity to sequence changes.

These metrics enable quantitative comparison of RBP binding properties across
different experimental conditions or between different RBPs.

## Reference

For a detailed description of these metrics and their biological applications, 
please refer to:

> Yi S, Singh SS, Ye X, Krishna R, Jankowsky E, Luna JM. (2025). 
> *Inherent Specificity and Mutational Sensitivity as Quantitative Metrics for RBP Binding.* 
> bioRxiv. https://www.biorxiv.org/content/10.1101/2025.03.28.646018v2

## Sample Data

The sample data included in this package were obtained from the ENCODE portal 
(Luo et al., 2020).

### RBNS Data

RNA Bind-n-Seq (RBNS) datasets provide in vitro RNA-RBP affinity measurements. 
For each K-mer enrichment dataset, the RBP concentration generating the highest 
R-value was selected as the representative sample. The collected RBNS R scores 
were feature-scaled to [1, e] followed by natural log transformation to normalize 
to a 0-to-1 scale.

### eCLIP Data

eCLIP datasets provide in vivo RBP binding site information. Peak coordinates 
were extended 25 nucleotides upstream from the 5'-end. Motif enrichment was 
calculated by counting K-mer appearances across extended peaks and subtracting 
background counts from randomly shifted genomic regions (up to 500 nt away). 
Background generation was repeated 100 times for robust averaging.

> Luo Y, Hitz BC, Gabdank I, et al. (2020). New developments on the Encyclopedia 
> of DNA Elements (ENCODE) data portal. *Nucleic Acids Res.* 48, D882–D889.

# Setup

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RBPSpecificity")
```

```{r load-package}
library(RBPSpecificity)
library(ggplot2)
```

# In Vitro Specificity from Pre-computed Enrichment Data (RBNS)

This workflow demonstrates how to calculate IS and VS from pre-computed 
enrichment scores, such as those from RBNS (RNA Bind-n-Seq) experiments.

## Loading RBNS Data

The sample data contains normalized 5-mer enrichment scores for four RBPs:
HNRNPC, PCBP2 (structural context independent), RBFOX2, and EIF4G2 
(structural context dependent).

```{r load-rbns}
# Load sample RBNS data from package
rbns_file <- system.file("extdata", "RBNS_normalized_5mer.csv", 
                          package = "RBPSpecificity")
rbns_data <- read.csv(rbns_file, stringsAsFactors = FALSE)

# Preview the data
head(rbns_data)
```

## Calculating IS and VS for All Four RBPs

The package expects a data.frame with `MOTIF` and `Score` columns:

```{r calc-rbns-metrics}
rbps <- c("HNRNPC", "PCBP2", "RBFOX2", "EIF4G2")

rbns_metrics <- list()
for (rbp in rbps) {
  rbns_enrichment <- data.frame(
    MOTIF = rbns_data$Motif,
    Score = rbns_data[[rbp]]
  )
  is_val <- returnIS(rbns_enrichment)
  vs_val <- returnVS(rbns_enrichment, output_type = "number")
  rbns_metrics[[rbp]] <- list(
    enrichment = rbns_enrichment,
    IS = is_val,
    VS = vs_val
  )
}

# Display in vitro metrics
invitro_summary <- data.frame(
  RBP = rbps,
  IS = sapply(rbns_metrics, function(x) round(x$IS, 2)),
  VS = sapply(rbns_metrics, function(x) round(x$VS, 4))
)
print(invitro_summary)
```

## Visualization

### IS Distribution Plots

The `plotIS()` function displays the distribution of k-mer scores, highlighting 
the top motif and its IS value:

```{r plot-is-rbns, fig.cap="Distribution of k-mer scores from RBNS data for four RBPs."}
plotIS(rbns_metrics[["HNRNPC"]]$enrichment)
plotIS(rbns_metrics[["PCBP2"]]$enrichment)
plotIS(rbns_metrics[["RBFOX2"]]$enrichment)
plotIS(rbns_metrics[["EIF4G2"]]$enrichment)
```

### VS Profile Plots

The `plotVS()` function visualizes how sensitive binding is to mutations at each 
position:

```{r plot-vs-rbns, fig.cap="Variation sensitivity profiles from RBNS data for four RBPs."}
plotVS(rbns_metrics[["HNRNPC"]]$enrichment)
plotVS(rbns_metrics[["PCBP2"]]$enrichment)
plotVS(rbns_metrics[["RBFOX2"]]$enrichment)
plotVS(rbns_metrics[["EIF4G2"]]$enrichment)
```

# Cellular Specificity from De Novo Enrichment (eCLIP)

This workflow demonstrates how to calculate enrichment scores directly from 
peak data using the `motifEnrichment()` function.

> **Note**: While this example uses eCLIP-derived peaks, `motifEnrichment()` 
> works with any peak data in BED format (e.g., iCLIP, PAR-CLIP, or other 
> ChIP-seq derived peaks).

> **Requirement**: This workflow requires `BSgenome.Hsapiens.UCSC.hg38` to be installed.

When calculating enrichment from cellular peak data, the resulting metrics are 
referred to as **Cellular Specificity (CS)** and **Cellular Variation 
Sensitivity (CVS)**, as they reflect in vivo binding behavior rather than 
intrinsic biochemical affinity.

## Loading Peak Data

```{r load-eclip, eval = requireNamespace("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE)}
# Load eCLIP peaks (10-column narrowPeak format) from package
bed_cols <- c("chr", "start", "end", "name", "score", "strand",
              "signalValue", "pValue", "qValue", "peak")

rbp_names <- c("HNRNPC", "PCBP2", "RBFOX2", "EIF4G2")
cell_lines <- c("K562", "HepG2", "K562", "K562")

peak_data <- list()
for (i in seq_along(rbp_names)) {
  bed_file <- system.file("extdata",
    sprintf("ENCODE_eCLIP_%s_%s_narrowPeak.bed", rbp_names[i], cell_lines[i]),
    package = "RBPSpecificity")
  peaks <- read.table(bed_file, header = FALSE, sep = "\t")
  colnames(peaks) <- bed_cols
  peak_data[[rbp_names[i]]] <- peaks
  message("Loaded ", nrow(peaks), " ", rbp_names[i], " peaks.")
}
```

## Running motifEnrichment Pipeline

In addition to standard input variables, `motifEnrichment()` provides several 
parameters for customization:

| Parameter | Description |
|-----------|-------------|
| `extension` | Numeric vector of length 2: `c(five_prime, three_prime)` indicating shifting of 5' and 3' ends. Positive extends, negative trims. |
| `method` | Enrichment model: `"anr"` (default, any number of repetitions, conditional Binomial test) or `"zoops"` (zero-or-one occurrence per sequence, Hypergeometric test) |
| `bkg_iter` | Number of background iterations for averaging (higher = more robust) |
| `bkg_min_dist` | Minimum distance (bp) to shift peaks when generating background regions |
| `bkg_max_dist` | Maximum distance (bp) for shifting peaks in background generation |
| `scramble_bkg` | Logical, scramble background sequences to control for nucleotide composition (default: FALSE) |
| `nucleic_acid_type`| Character string, `"DNA"` or `"RNA"` (default: `"DNA"`) |

ZOOPS method can also be used instead. Note that ANR and ZOOPS calculation can 
result in different motif enrichment values and thus different specificity and 
sensitivity calculations. For more information on ANR vs. ZOOPS, please check 
the GitHub README and Further Reading section. In addition, the scramble 
background option can be turned on (`scramble_bkg = TRUE`) to control for 
nucleotide composition.

```{r run-enrichment, eval = requireNamespace("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE)}
# NOTE: This requires BSgenome.Hsapiens.UCSC.hg38

cellular_results <- list()
for (rbp in rbp_names) {
  message("Running motifEnrichment for ", rbp, "...")
  enrichment <- motifEnrichment(
    coordinates = peak_data[[rbp]],
    species_or_build = "hg38",
    K = 5,
    extension = c(25, 0),
    method = "anr",
    bkg_iter = 100,
    scramble_bkg = FALSE
  )
  
  cs_val <- returnIS(enrichment)
  cvs_val <- returnVS(enrichment, output_type = "number")
  
  cellular_results[[rbp]] <- list(
    enrichment = enrichment,
    CS = cs_val,
    CVS = cvs_val
  )
}

# Display cellular metrics
cellular_summary <- data.frame(
  RBP = rbp_names,
  CS = sapply(cellular_results, function(x) round(x$CS, 2)),
  CVS = sapply(cellular_results, function(x) round(x$CVS, 4))
)
print(cellular_summary)
```

## Visualization

The same `plotIS()` and `plotVS()` functions used for RBNS data can be applied
to visualize cellular specificity (CS) and cellular variation sensitivity (CVS)
from eCLIP enrichment results:

```{r eclip-plots, eval = requireNamespace("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE)}
# CS Distribution plots
plotIS(cellular_results[["HNRNPC"]]$enrichment)
plotIS(cellular_results[["PCBP2"]]$enrichment)
plotIS(cellular_results[["RBFOX2"]]$enrichment)
plotIS(cellular_results[["EIF4G2"]]$enrichment)

# CVS Profile plots
plotVS(cellular_results[["HNRNPC"]]$enrichment)
plotVS(cellular_results[["PCBP2"]]$enrichment)
plotVS(cellular_results[["RBFOX2"]]$enrichment)
plotVS(cellular_results[["EIF4G2"]]$enrichment)
```

# Comparing In Vitro (RBNS) vs. Cellular (eCLIP) Specificity

A key application is comparing **inherent specificity** (IS/VS, from RBNS) with 
**cellular specificity** (CS/CVS, from eCLIP). RBPs that bind directly to 
single-stranded sequence motifs (e.g., HNRNPC, PCBP2) are expected to show 
correlated IS/CS and VS/CVS values. RBPs whose binding depends on structural 
context (e.g., RBFOX2, EIF4G2) may show divergence between in vitro and 
cellular metrics.

```{r comparison, eval = requireNamespace("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE)}
comparison <- data.frame(
  RBP = rbp_names,
  Structural_Context = c("Independent", "Independent", "Dependent", "Dependent"),
  In_Vitro_IS = sapply(rbns_metrics[rbp_names], function(x) round(x$IS, 2)),
  Cellular_CS = sapply(cellular_results, function(x) round(x$CS, 2)),
  In_Vitro_VS = sapply(rbns_metrics[rbp_names], function(x) round(x$VS, 4)),
  Cellular_CVS = sapply(cellular_results, function(x) round(x$CVS, 4))
)
print(comparison)
```

# Session Info

```{r session}
sessionInfo()
```
